---
title: "Macroeconomic Determinants of Supplier Payment Times"
output:
  html_document:
    toc: yes
    toc_depth: 3
author: "Abigail Gracias"
---


# Loading Libraries

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message = TRUE, eval = TRUE)
library(tidyverse)
library(plm)
library(car)
library(emmeans)
library(vtable)
library(psych)
library(lmtest)
```

# Read the Data

```{r}

#UK bank rates 2017 - 2024
uk_bank_rate = read_csv('uk_bank_rate.csv')

#UK company payment time
payment_times = read_csv('payment-practices.csv')

#UK GDP 2017 - 2024
uk_gdp = read_csv('uk_gdp.csv')

#UK unemployment rate 2017 - 2024
uk_unemp_rate = read_csv('uk_unemp_rate.csv')

#UK inflation rate 2-17 - 2024
uk_inflation = read_csv('uk_inflation_rate.csv')

#UK business confidence index 2017 - 2024
uk_BCI = read_csv('uk_BCI.csv')

#UK interest rates 2017-2024
uk_interest_rate = read_csv('uk_interest_rate.csv')

#UK company financials data
company_fdata = read_csv('company_FAME.csv') 

```
# Data Preparation

### Convert company financial data from wide to long

```{r}

#Extracting each wide variable set to separate temp dataframe

company_assets <- company_fdata %>% 
  select(`Company name`,`total_assets 2024`,`total_assets 2023`,`total_assets 2022`,`total_assets 2021`,`total_assets 2020`,`total_assets 2019`,`total_assets 2018`,`total_assets 2017`) %>%
  rename('company_name' = `Company name`,
         '2024' = `total_assets 2024`,
         '2023' = `total_assets 2023`,
         '2022' = `total_assets 2022`,
         '2021' = `total_assets 2021`,
         '2020' = `total_assets 2020`,
         '2019' = `total_assets 2019`,
         '2018' = `total_assets 2018`,
         '2017' = `total_assets 2017`) %>%
  gather( year, total_assets, '2024':'2017') %>%
  subset(year != '2024') 

company_assets <- company_assets %>% distinct(company_name, year, .keep_all = TRUE)
  
company_equity <- company_fdata %>% 
  select(`Company name`,`equity2024`,`equity2023`,`equity2022`,`equity2021`,`equity2020`,`equity2019`,`equity2018`,`equity2017`) %>%
  rename('company_name' = `Company name`,
         '2024' = `equity2024`,
         '2023' = `equity2023`,
         '2022' = `equity2022`,
         '2021' = `equity2021`,
         '2020' = `equity2020`,
         '2019' = `equity2019`,
         '2018' = `equity2018`,
         '2017' = `equity2017`) %>%
  gather( year, equity, '2024':'2017') %>%
  subset(year != '2024' ) 

company_equity <- company_equity %>% distinct(company_name, year, .keep_all = TRUE)
  
company_liability <- company_fdata %>% 
  select(`Company name`,`liabilities2024`,`liabilities2023`,`liabilities2022`,`liabilities2021`,`liabilities2020`,`liabilities2019`,`liabilities2018`,`liabilities2017`) %>%
  rename('company_name' = `Company name`,
         '2024' = `liabilities2024`,
         '2023' = `liabilities2023`,
         '2022' = `liabilities2022`,
         '2021' = `liabilities2021`,
         '2020' = `liabilities2020`,
         '2019' = `liabilities2019`,
         '2018' = `liabilities2018`,
         '2017' = `liabilities2017`) %>%
  gather( year, liabilities, '2024':'2017') %>%
  subset(year != '2024' )

company_liability <- company_liability %>% distinct(company_name, year, .keep_all = TRUE)
  
company_ROTA <- company_fdata %>% 
  select(`Company name`,`ROTA 2024`,`ROTA 2023`,`ROTA 2022`,`ROTA 2021`,`ROTA 2020`,`ROTA 2019`,`ROTA 2018`,`ROTA 2017`) %>%
  rename('company_name' = `Company name`,
         '2024' = `ROTA 2024`,
         '2023' = `ROTA 2023`,
         '2022' = `ROTA 2022`,
         '2021' = `ROTA 2021`,
         '2020' = `ROTA 2020`,
         '2019' = `ROTA 2019`,
         '2018' = `ROTA 2018`,
         '2017' = `ROTA 2017`) %>%
  gather( year, ROTA, '2024':'2017') %>%
  subset(year != '2024' )

company_ROTA <- company_ROTA %>% distinct(company_name, year, .keep_all = TRUE)
  
company_liquidity_ratio <- company_fdata %>% 
  select(`Company name`,`liquidity_ratio 2024`,`liquidity_ratio 2023`,`liquidity_ratio 2022`,`liquidity_ratio 2021`,`liquidity_ratio 2020`,`liquidity_ratio 2019`,`liquidity_ratio 2018`,`liquidity_ratio 2017`) %>%
  rename('company_name' = `Company name`,
         '2024' = `liquidity_ratio 2024`,
         '2023' = `liquidity_ratio 2023`,
         '2022' = `liquidity_ratio 2022`,
         '2021' = `liquidity_ratio 2021`,
         '2020' = `liquidity_ratio 2020`,
         '2019' = `liquidity_ratio 2019`,
         '2018' = `liquidity_ratio 2018`,
         '2017' = `liquidity_ratio 2017`) %>%
  gather( year, liquidity_ratio, '2024':'2017') %>%
  subset(year != '2024' )

company_liquidity_ratio <- company_liquidity_ratio %>% distinct(company_name, year, .keep_all = TRUE)
  
company_work_cap <- company_fdata %>% 
  select(`Company name`,`working_capital 2024`,`working_capital 2023`,`working_capital 2022`,`working_capital 2021`,`working_capital 2020`,`working_capital 2019`,`working_capital 2018`,`working_capital 2017`) %>%
  rename('company_name' = `Company name`,
         '2024' = `working_capital 2024`,
         '2023' = `working_capital 2023`,
         '2022' = `working_capital 2022`,
         '2021' = `working_capital 2021`,
         '2020' = `working_capital 2020`,
         '2019' = `working_capital 2019`,
         '2018' = `working_capital 2018`,
         '2017' = `working_capital 2017`) %>%
  gather( year, working_capital, '2024':'2017') %>%
  subset(year != '2024' )

company_work_cap <- company_work_cap %>% distinct(company_name, year, .keep_all = TRUE)

  
company_credit_pay <- company_fdata %>% 
  select(`Company name`,`creditors_payment 2024`,`creditors_payment 2023`,`creditors_payment 2022`,`creditors_payment 2021`,`creditors_payment 2020`,`creditors_payment 2019`,`creditors_payment 2018`,`creditors_payment 2017`) %>%
  rename('company_name' = `Company name`,
         '2024' = `creditors_payment 2024`,
         '2023' = `creditors_payment 2023`,
         '2022' = `creditors_payment 2022`,
         '2021' = `creditors_payment 2021`,
         '2020' = `creditors_payment 2020`,
         '2019' = `creditors_payment 2019`,
         '2018' = `creditors_payment 2018`,
         '2017' = `creditors_payment 2017`) %>%
  gather( year, creditors_payment, '2024':'2017') %>%
  subset(year != '2024' )

company_credit_pay <- company_credit_pay %>% distinct(company_name, year, .keep_all = TRUE)


company_fdata <- select(company_fdata, `Company name`, `BvD sector`) %>%
  rename(company_name = `Company name`,
         sector = `BvD sector`)

company_fdata <- company_fdata %>% distinct(company_name, .keep_all = TRUE)

```


### Prepare the transformed company data frame

```{r}

# Only keep data of companies having records for all years
companies_with_na <- rbind(company_assets[!complete.cases(company_assets),] %>% select(company_name), 
                 company_credit_pay[!complete.cases(company_credit_pay),] %>% select(company_name),
                 company_equity[!complete.cases(company_equity),] %>% select(company_name),
                 company_liability[!complete.cases(company_liability),] %>% select(company_name),
                 company_liquidity_ratio[!complete.cases(company_liquidity_ratio),] %>% select(company_name),
                 company_ROTA[!complete.cases(company_ROTA),] %>% select(company_name),
                 company_work_cap[!complete.cases(company_work_cap),] %>% select(company_name))

companies_with_na <- companies_with_na %>% distinct(company_name, .keep_all = TRUE)

# Combining all company data and ensure the 'date' column is in Date format (first day of the year) 
company_data <- company_assets %>% 
  left_join( company_equity, by=c('company_name', 'year')) %>%
  left_join( company_liability, by=c('company_name', 'year')) %>%
  left_join( company_ROTA, by=c('company_name', 'year')) %>%
  left_join( company_work_cap, by=c('company_name', 'year')) %>%
  left_join( company_credit_pay, by=c('company_name', 'year')) %>%
  left_join( company_liquidity_ratio, by=c('company_name', 'year')) %>%
  left_join( company_fdata, by=c('company_name')) %>%
  mutate(date = ymd(paste0(year, "-01-01"))) %>%
  select(-year) 

company_data <- company_data %>%
  filter(!company_name %in% companies_with_na$company_name)

# Function to generate monthly dates for a given year
generate_monthly_dates <- function(year_date) {
  seq(from = year_date, 
      to = year_date + years(1) - months(1), 
      by = "month")
}

# Apply the function to each row to standardise the data
company_data <- company_data %>%
  rowwise() %>%
  mutate(month_sequence = list(generate_monthly_dates(date))) %>%
  unnest(month_sequence) %>%
  mutate(date = format(month_sequence, "%m/%Y")) %>%
  select(-month_sequence) %>% 
  mutate(date = my(date))


#Removing temp data frames
remove(company_assets)
remove(company_credit_pay)
remove(company_equity)
remove(company_liability)
remove(company_liquidity_ratio)
remove(company_ROTA)
remove(company_work_cap)

```



### Check for NA values from macro economic data

```{r}
sum(is.na(uk_bank_rate))

sum(is.na(uk_gdp))

sum(is.na(uk_unemp_rate))

sum(is.na(uk_inflation))

#replacing NA values with prev quarter's value
uk_BCI <-uk_BCI %>% fill(BCI)
sum(is.na(uk_BCI))

sum(is.na(uk_interest_rate))

```


### Formatting all dates to MM/YYY

```{r}
# Ensure date columns are in Date format yyyy-mm-dd
payment_times <- payment_times %>%
  mutate(
    Start_date = dmy(Start_date),
    End_date = dmy(End_date)
  )

uk_interest_rate <- uk_interest_rate %>%
  mutate(Date = dmy(Date),
         Date = format(Date, "%m/%Y"))

uk_gdp <- uk_gdp %>%
  mutate(year = yq(year)) %>%
  rename(date = year)


uk_unemp_rate <- uk_unemp_rate %>%
  mutate(month = dplyr::recode(month,
  JAN = 1,
  FEB = 2,
  MAR = 3,
  APR = 4,
  MAY = 5,
  JUN = 6,
  JUL = 7,
  AUG = 8,
  SEP = 9,
  OCT = 10,
  NOV = 11,
  DEC = 12)) %>%
  unite(col="updated_year", c(year,month)) %>%
  mutate(updated_year = ym(updated_year)) %>%
  rename(date = updated_year) %>%
  mutate(date = format(date, "%m/%Y"))

uk_inflation <- uk_inflation %>%
   mutate(month = dplyr::recode(month,
  JAN = 1,
  FEB = 2,
  MAR = 3,
  APR = 4,
  MAY = 5,
  JUN = 6,
  JUL = 7,
  AUG = 8,
  SEP = 9,
  OCT = 10,
  NOV = 11,
  DEC = 12)) %>%
  unite(col="updated_year", c(year,month)) %>%
  mutate(updated_year = ym(updated_year)) %>%
  rename(date = updated_year) %>%
  mutate(date = format(date, "%m/%Y"))

uk_BCI <- uk_BCI %>%
  unite(col = "updated_year", c(year,quarter)) %>%
  mutate(updated_year = yq(updated_year)) %>%
  rename(date = updated_year)

uk_bank_rate <- uk_bank_rate %>%
  mutate(
    Date = dmy(Date),
    month_year = format(Date, "%m/%Y")) %>%
  rename(bank_rate = `Bank Rate`)

```


### Standardizing all dates to be monthly

```{r}

# Function to generate sequence of months between start and end dates
generate_month_sequence_1 <- function(Start_date, End_date) {
  seq(from = floor_date(Start_date, "month"), 
      to = floor_date(End_date, "month"), 
      by = "month")
}

# Apply the function to each row 
payment_times <- payment_times %>%
  rowwise() %>%
  mutate(month_sequence = list(generate_month_sequence_1(Start_date, End_date))) %>%
  unnest(month_sequence) %>%
  mutate(date = format(month_sequence, "%m/%Y")) %>%
  select(-month_sequence, -Report_Id, - Start_date, - End_date)

#there are 10208 distinct companies originally in the dataset
n_distinct(payment_times$Company)

# extract all companies having NA values
na_rows <- payment_times[!complete.cases(payment_times),] %>% select(Company) %>% distinct(Company, .keep_all = TRUE)
payment_times <- payment_times %>% filter(!Company %in% na_rows$Company)

#Check for NA values
sum(is.na(payment_times))

#Filter out 2024 data
payment_times <- payment_times %>% filter(!grepl("2024",date))

#there are now 7979 distinct companies after removing companies with NA values


# bank rate transformed from daily to monthly by taking mean of each months data
uk_bank_rate <- uk_bank_rate %>%
  group_by(month_year) %>%
  summarise(bank_rate = mean(bank_rate)) %>%
  rename(date = month_year)
  

# Function to convert quarterly dates for the year
generate_month_sequence_3 <- function(year_quarter) {
  seq(from = year_quarter, by = "month", length.out = 3)
}

# Apply the function to each row and duplicate records
uk_BCI <- uk_BCI %>%
  rowwise() %>%
  mutate(month_sequence = list(generate_month_sequence_3(date))) %>%
  unnest(month_sequence) %>%
  mutate(date = format(month_sequence, "%m/%Y")) %>%
  select(-month_sequence)

uk_gdp <- uk_gdp %>%
  rowwise() %>%
  mutate(month_sequence = list(generate_month_sequence_3(date))) %>%
  unnest(month_sequence) %>%
  mutate(date = format(month_sequence, "%m/%Y")) %>%
  select(-month_sequence)

```


### Creating the master data set using payment_times

```{r}

# Left joining supplier payment time data to macroeconomic factors
master_data <- payment_times %>% 
  left_join( uk_bank_rate, by=c('date' = 'date')) %>%
  left_join( uk_BCI, by=c('date' = 'date')) %>%
  left_join( uk_gdp, by=c('date' = 'date')) %>%
  left_join( uk_inflation, by=c('date' = 'date')) %>%
  left_join( uk_unemp_rate, by=c('date' = 'date')) %>%
  left_join(uk_interest_rate, by=c('date' = 'Date')) %>%
  mutate(date = my(date))

# Left joining master data to company data (control variables data)
master_data <- master_data %>%
  left_join(company_data, by=c('Company' = 'company_name', 'date' = 'date')) %>%
  rename('interest_rate' = `interest rate`) %>%
  mutate('debt_to_equity' = liabilities/equity) %>%
  select(-liabilities, - equity)

# extract all companies having NA values
na_rows <- master_data[!complete.cases(master_data),] %>% select(Company) %>% distinct(Company, .keep_all = TRUE)

master_data_na_omit <- master_data %>% filter(!Company %in% na_rows$Company)

#Check for NA values
sum(is.na(master_data))

#Convert sector and Company to factor
master_data_na_omit$sector <- as.factor(master_data_na_omit$sector)
master_data_na_omit$Company <- as.factor(master_data_na_omit$Company)

#Remove duplicates
master_data_distinct <- master_data_na_omit %>% distinct(Company, date, .keep_all = TRUE)

#write final dataframe to CSV file
write.csv(master_data_distinct, file = "master_data.csv")

```


### Renaming attributes for clarity
```{r}
master_data_distinct <- master_data_distinct %>%
  rename("Supplier_Payment_Time" = "Average_time_to_pay" ,
         "Bank_Rate" = "bank_rate" ,
         "Business_Confidence_Index" = "BCI" ,
         "GDP_Growth_Rate" = "GDP" ,
         "Inflation_Rate" = "inflation_rate",
         "Unemployment_Rate" = "unemployment_rate" ,
         "Interest_Rate" = "interest_rate" ,
         "Firm_Size" = "total_assets" ,
         "Profitability" = "ROTA" ,
         "Access_to_Credit" = "debt_to_equity" ,
         "Industry" = "sector")

#check dataframe structure
str(master_data_distinct)

```



# Exploratory Data Analysis


### Summary statistics

```{r}
# View summary statistics of the master data
st(master_data_distinct, vars= c("Supplier_Payment_Time", "Bank_Rate", "Business_Confidence_Index", "GDP_Growth_Rate", "Inflation_Rate", "Unemployment_Rate", "Interest_Rate", "Firm_Size", "Profitability","Access_to_Credit"))

```


### Visualisation of Macro-Economic indicators (independent variables)

```{r}
master_data_distinct %>%
  ggplot(aes(date, y=Interest_Rate)) + geom_line(alpha=0.5) + labs(x="Time / Month", y="Interest rate")

master_data_distinct %>%
  ggplot(aes(date, y=Bank_Rate)) + geom_line(alpha=0.5) + labs(x="Time / Month", y="Bank rate")

master_data_distinct %>%
  ggplot(aes(date, y=Business_Confidence_Index)) + geom_line(alpha=0.5) + labs(x="Time / Month", y="Business Confidence Index")

master_data_distinct %>%
  ggplot(aes(date, y=Inflation_Rate)) + geom_line(alpha=0.5) + labs(x="Time / Month", y="Inflation rate")

master_data_distinct %>%
  ggplot(aes(date, y=Unemployment_Rate)) + geom_line(alpha=0.5) + labs(x="Time / Month", y="Unemployment rate")

master_data_distinct %>%
  ggplot(aes(date, y=GDP_Growth_Rate)) + geom_line(alpha=0.5) + labs(x="Time / Month", y="GDP Growth Rate")


```



### Visualisation of supplier payment time (dependent variable)

```{r}
master_data_distinct %>%
  group_by(Industry, date) %>%
  summarise(Supplier_Payment_Time = mean(Supplier_Payment_Time)) %>%
  ggplot(aes(x=date, y=Supplier_Payment_Time, col=Industry)) + geom_line(alpha=0.5) + scale_y_continuous(breaks=1:10) + labs(x="Time / Month", y="Supplier Payment Time", col="Industry") + geom_vline(xintercept=2.5, lty=3) +
  facet_wrap(~ Industry, nrow=3)

```


### Identifying Outliers

```{r}

master_data_distinct %>%
  ggplot(aes(date,y=Supplier_Payment_Time)) + geom_point(alpha=0.5)

#master_data_distinct %>% group_by(Supplier_Payment_Time) %>% summarise(COUNT = n())

# removing outliers
master_data_distinct <- filter(master_data_distinct, master_data_distinct$Supplier_Payment_Time<2000)

```



### Checking for multicolinearity

```{r}

# visual inspection
ggplot(master_data_distinct, aes(x=Interest_Rate, y=Bank_Rate)) + geom_point() + labs(x="Interest rate", y="Bank rate") + geom_vline(xintercept=2.5, lty=3)

# numerical measures

lowerCor(master_data_distinct[, c("Bank_Rate", "Business_Confidence_Index", "GDP_Growth_Rate", "Inflation_Rate", 
                                      "Unemployment_Rate", "Interest_Rate", 
                                      "Firm_Size", "Profitability", "Access_to_Credit")])

```
Bank rate removed due to high collinearity with Interest rate (0.99)


# Model Implementation

### Running a fixed effects model

```{r}
# Fit a Fixed Effects Model
remove(fixed_effects_model)

fixed_effects_model <- plm(Supplier_Payment_Time ~ Business_Confidence_Index+GDP_Growth_Rate+Inflation_Rate+Unemployment_Rate+Interest_Rate+Firm_Size+Profitability+Access_to_Credit+Industry, 
                           data = master_data_distinct, 
                           index = c("Company","date"), 
                           model = "within")

summary(fixed_effects_model)

res <- summary(fixed_effects_model)
print(res$coefficients)

```


### Checking FE model for Autocorelation and Heteroskedasticity

```{r}
#testing for autocorelation
bg_test <- bgtest(fixed_effects_model, order = 1)
print(bg_test)

#testing for heteroskedasticity
bp_test <- bptest(fixed_effects_model)
print(bp_test)
```

For Breusch-Godfrey Test for Autocorrelation p-value < 0.05, null hypothesis is rejected, meaning there is evidence of autocorrelation

For Breusch-Pagan Test for Heteroskedasticity p-value < 0.05, null hypothesis is rejected, indicating the presence of heteroskedasticity


```{r}

#Adjusting for heteroskedasticity using robust standard errors
coeftest(fixed_effects_model, vcovHC(fixed_effects_model, type = "HC1"))

```

### Running a Random Effects model

```{r}

# Fit a random effects model
random_effects_model <- plm(Supplier_Payment_Time ~ Business_Confidence_Index+GDP_Growth_Rate+Unemployment_Rate+Inflation_Rate+Interest_Rate+Firm_Size+Profitability+Access_to_Credit+Industry, 
                           data = master_data_distinct, 
                           index = c("Company","date"), 
                           model = "random")

summary(random_effects_model)

```

### Comparing fit of both models through Hausman test

```{r}

hausman_test <- phtest(fixed_effects_model,random_effects_model)
print(hausman_test)

```
 
 Result shows p-value < 0.05, null hypothesis is rejected, the fixed effects model is more appropriate.


